A Type System for Safe and Scalable Parallelism using Logical Regions

### 2012-07-11 doc/popl2013/intro.tex
s/permission/privilege/g

In Legion data is organized in a hierarchy of {\em regions}
and subregions while computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region permissions} specifying 
operations a task may perform on a region argument (read,
write, or reduce) and  {\em region coherence} annotations that
express what other tasks may do concurrently with
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the permissions and coherence system and use these two
results to prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then the any descendent of $t_1$ in the task hierarchy is non-interfering
with any descendent of $t_2$.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in communicating in the target
class of machines.

@@

Our own experience with writing high-performance applications in both 
Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, fairly static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple partitions simultaneously of
the same data, providing different views on to that data. Thus, in Legion the 
handling of regions and partitioning of regions is much more dynamic than in 
previous designs.  In particular,
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

% Moving this to the circuit example section since it fits better there
% and some of this is redundant
%Legion's execution model is that by default, a program's semantics is
%its meaning as a sequential program, which we refer to as the {\em
%program order} execution.  While the
%implementation is not our focus here, if two functions use
%disjoint data (because all of the regions they use are disjoint), then
%Legion may execute them simultaneously.  Also like DPJ,
%Legion has a system of {\em permissions} ({\em read}, {\em write}, and
%{\em reduce}) on regions and an orthogonal system of region {\em
%coherence} modes ({\em excl}, {\em atomic}, and {\em simultaneous})
%that increase the number of situations in which functions can be
%executed in parallel.

The important difference bewteen Legion and the more static approaches
is that Legion allows aliasing, but does so in a structured way that minimizes
runtime overheads.  The key observation is that
static alias analysis is hard, but dynamic analysis is very easy
and relatively cheap
when done at the granularity of regions instead of individual memory
locations.  Thus, Legion falls between fully static systems, such as
Sequoia and DPJ, and fully dynamic approaches such as transactional
memory.  Legion still has a significant static component; in
particular, the safety of permissions is checked statically and region
pointer dereferences are checked statically.  The dynamic alias tests
for independence are done at the granularity of regions and one check 
on a region
often serves to prove the independence of many subsequent uses of that region.
In contrast, because transactional memory has no mechanism
for grouping heap data into larger collections, it must test the
overlap between two sets of data on each individual memory access, which is
significantly more expensive.

We begin in Section~\ref{sec:example} with an example program that illustrates
a typical style of programming in Legion and motivates the need for multiple, dynamically computed partitions of
a region.  We define Core Legion, a small language suitable for our formal development, in Section~\ref{sec:legioncore}.
The next four sections each state and prove one of our main results and contributions:
\begin{itemize}
\item We prove the soundness of Legion's permissions system (Section~\ref{sec:soundness}).

\item We use the soundness of permissions to prove the soundness of Legion's region coherence modes (Section~\ref{sec:coherence}).

@@

\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
We also show that the cost of checking aliasing on region permissions is low, showing that a much more expressive
and dynamic langauge is not incompatible with high performance.

>>>>

In Legion data is organized in a hierarchy of {\em regions}
and subregions while computation is organized in a hierarchy of {\em
tasks} and subtasks operating on regions.  Regions and tasks interact
through a static system of {\em region privileges} specifying 
operations a task may perform on a region argument (read,
write, or reduce) and  {\em region coherence} annotations that
express what other tasks may do concurrently with
the region (exclusive, atomic, or simultaneous).  We prove the
soundness of the privileges and coherence system and use these two
results to prove a third result: if two {\em siblings} (tasks that
have the same immediate parent task) $t_1$ and $t_2$ are 
{\em non-interfering} (have no ordering requirements for correctness), 
then the any descendent of $t_1$ in the task hierarchy is non-interfering
with any descendent of $t_2$.  This theorem is the basis for correct distributed
task scheduling in a Legion implementation, which is crucial for
performance; a centralized scheduler would be a bottleneck
because of the huge latencies involved in communicating in the target
class of machines.

@@

Our own experience with writing high-performance applications in both 
Sequoia and the current industry standard mix of MPI, shared-memory
threads, and CUDA, has taught us that for many practical situations a
single, static partitioning is insufficient.  In practice, the
best way to partition data is often a function of the data
itself---i.e., the partitions need to be dynamically computed.
Furthermore, it is useful to allow multiple simultaneous partitions of
the same data, providing different views on to that data. Thus, in Legion the 
handling of regions and partitioning of regions is much more dynamic than in 
previous designs.  In particular,
\begin{itemize}
\item  logical regions are first-class values in Legion
and may be dynamically allocated and stored in data structures;

@@

% Moving this to the circuit example section since it fits better there
% and some of this is redundant
%Legion's execution model is that by default, a program's semantics is
%its meaning as a sequential program, which we refer to as the {\em
%program order} execution.  While the
%implementation is not our focus here, if two functions use
%disjoint data (because all of the regions they use are disjoint), then
%Legion may execute them simultaneously.  Also like DPJ,
%Legion has a system of {\em privileges} ({\em read}, {\em write}, and
%{\em reduce}) on regions and an orthogonal system of region {\em
%coherence} modes ({\em excl}, {\em atomic}, and {\em simultaneous})
%that increase the number of situations in which functions can be
%executed in parallel.

The important difference bewteen Legion and the more static approaches
is that Legion allows aliasing, but does so in a structured way that minimizes
runtime overheads.  The key observation is that
static alias analysis is hard, but dynamic analysis is very easy
and relatively cheap
when done at the granularity of regions instead of individual memory
locations.  Thus, Legion falls between fully static systems, such as
Sequoia and DPJ, and fully dynamic approaches such as transactional
memory.  Legion still has a significant static component; in
particular, the required privileges for function calls is checked statically
and region
pointer dereferences are checked statically.  The dynamic alias tests
for independence are done at the granularity of regions and one check 
on a region
often serves to prove the independence of many subsequent uses of that region.
In contrast, because transactional memory has no mechanism
for grouping heap data into larger collections, it must test the
overlap between two sets of data on each individual memory access, which is
significantly more expensive.

We begin in Section~\ref{sec:example} with an example program that illustrates
a typical style of programming in Legion and motivates the need for multiple, dynamically computed partitions of
a region.  We define Core Legion, a small language suitable for our formal development, in Section~\ref{sec:legioncore}.
The next four sections each state and prove one of our main results and contributions:
\begin{itemize}
\item We prove the soundness of Legion's privileges system (Section~\ref{sec:soundness}).

\item We use the soundness of privileges to prove the soundness of Legion's region coherence modes (Section~\ref{sec:coherence}).

@@

\item We give experimental evidence for the Legion design.  On three real-world applications, we show that 
dynamic region pointer checks would be expensive, justifying checking this aspect of the type system statically.
We also show that the cost of checking for non-interference of region privileges is low, showing that a much more expressive
and dynamic langauge is not incompatible with high performance.
